---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# weda

  <!-- badges: start -->
  [![R-CMD-check](https://github.com/JustinCally/weda/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JustinCally/weda/actions/workflows/R-CMD-check.yaml)
  [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
  <!-- badges: end -->

`weda` (Wildlife Ecology Data Analytics) is an R package for users at the *Arthur Rylah Institute* to connect to and use the Wildlife Ecology database. The package has demonstrations on how to upload and data to the database (e.g. camera trap data). Functions in this package will allow users to format data in a standardised way, assess data quality and interact with the database.  

## Installation

You can install the development version of weda from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JustinCally/weda")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, eval = FALSE}
library(weda)
## basic example code
```

