---
title: "Connecting to the database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting to the database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = FALSE}
library(weda)
library(DBI)
library(dplyr)
library(dbplyr)
library(RPostgreSQL)
library(odbc)
library(keyring)
```

# Brief  

The Wildlife Ecology Section currently has an *Azure* virtual machine (VM) hosted through DEECA's ISD offerings. The VM is a linux system that has a *postgresql* database installed. This database has daily backups and is accessible through private endpoints (meaning users must have GoConnect/an azure VPN running). Access to the database can be granted by admins (Justin Cally) and will require a username and password to connect. Below we show how you can connect to the database from R/RStudio. Once connected, you are able to use `SQL` queries or translated `R` code (e.g. `dbplyr` R package) to interact with the database.    

# Step 1: Switch on VPN  

Make sure your Azure VPN (or GoConnect) in running  

# Step 2: Set keyring password if not set  

Set up keyring if not already set. Password for `psql_user` will be remembered in Rstudio keychain once you set it. Ask an admin for an account if you do not already have one.  

```{r keyring, eval = FALSE}
if(nrow(keyring::key_list("ari-dev-weda-psql-01")) == 0) {
keyring::key_set(service = "ari-dev-weda-psql-01", username = "psql_user")
}
```

# Step 3: Check capability  

Check if you have the correct odbc driver

```{r postgres, eval = FALSE}
# Check odbc driver exists
if(!("PostgreSQL Driver" %in% odbc::odbcListDrivers()$name)) {
  stop("Install the postgres driver to connect to database: https://odbc.postgresql.org/")
}
```

# Step 4: Connect to database  

Make a connection object in `R`

```{r connect, eval = FALSE}
# Note that the DB is currently configured to turn on at 7am and turn off at 7pm to reduce costs
# ODBC Connection : more interactive in rstudio:
con <- weda_connect(password = keyring::key_get(service = "ari-dev-weda-psql-01", 
                                                username = "psql_user"))

#### Option 2: Connect using manual connection string ####
con <- RPostgreSQL::dbConnect(odbc::odbc(),
                              Driver = "PostgreSQL Driver",
                              Server = '10.110.7.201',
                              Database = 'ari-dev-weda-psql-01',
                              UID = "psql_user",
                              PWD = keyring::key_get(service = "ari-dev-weda-psql-01", username = "psql_user"),
                              Port = 5432,
                              BoolsAsChar = 'No',
                              sslmode = 'require',
                              maxvarcharsize = 0) # issue with binary representation of sf cols
```

# Step 5: View data   

Interact with the database with the 'Connections' pane in Rstudio or use code such as the example below to read in data:  

```{r view, eval = FALSE}
#### Read and write tables ####
DBI::dbWriteTable(con, Id(schema = "test", table = "mtcars"), mtcars)

# Use lazy evaluation to download a sample of mtcars data
db_mtcars <- tbl(con, in_schema("test", "mtcars")) %>%
  filter(cyl == 6) %>%
  collect()
```

# Additional Resources  

You can follow a further tutorial of postgres database functionality here:  

```{r, file=system.file('DBConnect.R', package = "weda"), eval = FALSE}
```

