---
title: "Uploading transect records to database"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{transect-upload}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = TRUE, results='hide', message=FALSE, warning=FALSE}
library(weda)
library(dplyr)
library(sf)
library(kableExtra)
```

# Brief

We need a process that takes (i) raw proofsafe survey records, (ii) GPS transect lines in GeoJSON format, and (iii) project information, and uploads them to the database in a consistent manner that allows transect survey data across projects to be stored and analysed together.

Transect surveys at ARI follow a **double-observer distance-sampling** design, where two observers walk the same transect and independently record detections. Surveys are conducted either during the day (e.g. koala surveys) or at night via spotlight (e.g. greater glider surveys). Raw data is captured in the field using [Proofsafe](https://www.proofsafe.com.au/) forms and exported as CSV files.

# Input

There are three inputs to this process:

1. A **proofsafe records table** — the raw CSV export from Proofsafe containing animal detections along transects
2. A **GeoJSON file of GPS transect lines** — spatial line features with `SiteID` and `Transect` columns linking back to the records
3. A **project information table** — a single-row CSV with metadata about the project (e.g. project name, target species, survey design)

See the [appendix](#appendix) for the full list of columns expected in each uploaded table.

# Transect database model

Once this process is completed, data will be uploaded into the `raw` layer of the `transect` schema on the database. Automatic database views will then process this raw data into the `curated` and `processed` layers.

**Raw layer** — all records as uploaded, including any re-uploads (de-duplication handled by curated views):

- `raw_transect_records` — individual animal detections
- `raw_transects` — transect deployment information (with spatial geometry)
- `raw_project_information` — project metadata

**Curated layer** — duplicates removed, most data fields retained:

- `curated_transect_records` — most recent records for each unique detection
- `curated_transects` — most recent entry for each transect deployment
- `curated_project_information` — most recent entry for each project

**Processed layer** — derived outputs ready for analysis:

- `processed_transect_presence_absence` — presence/absence of each species at each transect × site × iteration combination, with spatial geometry attached

# Process

Uploading transect data involves eight steps. The easiest way to complete these steps is through the interactive Shiny app described below. Alternatively, the steps can be run directly in R using the functions described in [Upload using R code](#upload-using-r-code).

**Step 1** — Import the proofsafe records CSV
**Step 2** — Import the GeoJSON transect lines
**Step 3** — Import the project information CSV
**Step 4** — Format proofsafe data to database format (applies the relevant `*_proofsafe_format()` function)
**Step 5** — Standardise species names to VBA taxonomy
**Step 6** — Inspect a map of records and transects
**Step 7** — Run data quality checks (all checks must pass before upload)
**Step 8** — Upload to database

# Upload data using the Shiny app

The easiest way to upload transect data is through the hosted Shiny app at:

**<https://arisci.shinyapps.io/transect-app/>**

*Note: the app requires VPN access and login credentials.*

The app walks you through all eight steps in the sidebar. Download the example data from the "Download Example Data" button in the app to see the expected file formats before preparing your own data.

The app currently uses the `region_gg_proofsafe_format()` function in Step 4. If you need to upload **koala** survey data or standard (non-regional) **greater glider** data, use the R code workflow below with `koala_proofsafe_format()` or `gg_proofsafe_format()` respectively.

You can also run the app locally from R if needed. Set up a database connection and launch the app:

*Note: to learn more about database connections, see `vignette('database-connect')`*

```{r launchapp, eval = FALSE}
# Make sure VPN is running
con <- weda::weda_connect(password = keyring::key_get(service = "ari-dev-weda-psql-01",
                                                      username = "psql_user"))
# Launch the transect app
shiny::runApp(system.file("app/transect-app.R", package = "weda"))
```

# Upload using R code

The following steps replicate what the Shiny app does, but in R code. This is useful when the app's default format function does not match your survey type, or when you want to script a repeatable upload workflow.

Example data files used below are bundled with the package. The regional greater glider example uses a raw proofsafe export (`region_gg_records.csv`) alongside its GPS transect lines and project metadata:

```{r examplefiles}
proofsafe_path <- system.file("dummydata/transectdata/region_gg_records.csv", package = "weda")
transects_path <- system.file("dummydata/transectdata/region_gg_transects.geojson", package = "weda")
project_path   <- system.file("dummydata/transectdata/region_gg_project.csv", package = "weda")
```

## Step 1 — Load proofsafe records

Load the raw proofsafe CSV export. This is the direct output from exporting a Proofsafe form — no pre-processing is needed at this stage.

```{r loadrecords}
raw_proofsafe <- readr::read_csv(proofsafe_path, show_col_types = FALSE)

raw_proofsafe %>%
  head(5) %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%")
```

## Step 2 — Load GPS transect lines

The transect geometry must be provided as a GeoJSON file containing line features. The GeoJSON must include at least a `SiteID` and `Transect` column so that spatial lines can be joined to the survey records.

```{r loadtransects}
gps_transects <- sf::st_read(transects_path, quiet = TRUE)

gps_transects %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%")
```

## Step 3 — Create project information

Alongside the survey data, a single-row table is required to record project-level metadata. You can load this from a CSV (as done in the app) or construct it directly in R. Check that your `ProjectShortName` is not already in use by another project with `check_unique_project()`.

```{r projectdata}
project_information <- readr::read_csv(project_path, show_col_types = FALSE)

project_information %>%
  kbl() %>%
  kable_styling() %>%
  scroll_box(width = "100%")
```

The required fields are:

| Column | Description |
|---|---|
| `ProjectName` | Full descriptive project name |
| `ProjectShortName` | Short unique identifier used as a key in the database |
| `DistanceSampling` | Logical — whether the survey used distance sampling |
| `TerrestrialArboreal` | `"Terrestrial"` or `"Arboreal"` |
| `AllSpeciesTagged` | Logical — whether all encountered species were recorded |
| `DistanceForAllSpecies` | Logical — whether distances were recorded for all species |
| `DiurnalNocturnal` | `"Diurnal"` or `"Nocturnal"` |
| `ProjectDescription` | Brief description of the project objectives |
| `ProjectLeader` | Name of the project lead |

## Step 4 — Format proofsafe data to database format

This is the key transformation step. The appropriate `*_proofsafe_format()` function converts the raw proofsafe CSV and GPS transect lines into the standardised database format, calculating derived fields such as perpendicular distances and projected animal locations.

Choose the function that matches your survey type:

| Survey type | Function |
|---|---|
| Koala (diurnal double-observer) | `koala_proofsafe_format()` |
| Greater glider spotlight (standard) | `gg_proofsafe_format()` |
| Greater glider spotlight (regional) | `region_gg_proofsafe_format()` |

For this example we use `region_gg_proofsafe_format()`:

```{r formatdata}
formatted_data <- region_gg_proofsafe_format(
  proofsafe          = raw_proofsafe,
  gps_transects      = gps_transects,
  Iteration          = 1L,
  MaxTruncationDistance = 100
)

# The output is a list with two elements:
names(formatted_data)
```

For koala surveys, the call is equivalent but uses `koala_proofsafe_format()`:

```{r formatkola, eval = FALSE}
formatted_data <- koala_proofsafe_format(
  proofsafe             = raw_proofsafe,
  gps_transects         = gps_transects,
  sp_filter             = "Koala",
  Iteration             = 1L,
  SurveyMethod          = "Diurnal double-observer distance-sampling",
  MaxTruncationDistance = 100
)
```

The returned list contains:

- `$records` — data.frame of formatted animal detections
- `$transects` — sf data.frame of transect deployments with spatial geometry

## Step 5 — Standardise species names

Species names in the database follow the Victorian Biodiversity Atlas (VBA) taxonomy, which requires both a `scientific_name` and a `common_name` column. Use `standardise_species_names()` to check and append the missing name type from the VBA lookup table (`weda::vba_name_conversions`).

First, check whether your species names match the VBA without modifying the data:

```{r checknames}
standardise_species_names(recordTable = formatted_data$records,
                          format      = "scientific",
                          speciesCol  = "Species",
                          return_data = FALSE)
```

If all names convert successfully, standardise the data:

```{r standardisenames}
formatted_records_std <- standardise_species_names(
  recordTable = formatted_data$records,
  format      = "scientific",
  speciesCol  = "Species",
  return_data = TRUE
)
```

If a name cannot be matched, you will need to correct it in the source data to align with VBA conventions. The full VBA lookup can be browsed with `View(weda::vba_name_conversions)`.

## Step 6 — Inspect a map of records and transects

Before running data quality checks, it is worth visually inspecting the data to confirm that animal records fall in plausible locations relative to the transect lines.

```{r visualise}
visualise_records(records   = formatted_records_std,
                  transects = formatted_data$transects)
```

The map shows:

- **Blue lines/polygons** — transect lines and their truncation-distance buffers
- **Red points** — projected animal locations
- **Black lines** — observer-to-animal sight-lines
- **Orange lines** — second-observer projections (where dual-observer data exists)

Records that fall outside the transect buffer can be inspected or removed using `filter_records_outside_transect_area()`.

## Step 7 — Run data quality checks

All data must pass the quality checks before upload. The `transect_dq()` function runs over 100 checks across the three tables using the `pointblank` package, covering column presence, data types, value ranges, coordinate bounds, and distance-sampling constraints.

```{r dataquality}
dq <- transect_dq(
  records             = formatted_records_std,
  transects           = formatted_data$transects,
  project_information = project_information
)

dq[[1]]  # records
dq[[2]]  # transects
dq[[3]]  # project information
```

Each output table shows:

- **W (Warning)** — yellow dot; data is usable but should be reviewed
- **S (Stop)** — red dot; data cannot be uploaded until the issue is resolved
- **N (Notify)** — informational flag for unusual but not invalid values

Upload will only proceed when there are no red dots. Use the **EXT** download button on any failing row to export the problematic records and trace them back to your source data.

To confirm all checks have passed programmatically:

```{r checkpassed}
all(sapply(dq, function(x) all(x[["validation_set"]][["all_passed"]])))
```

## Step 8 — Prepare and upload

Once all quality checks pass, prepare the data for upload. This step generates MD5 hash IDs for each record to prevent duplicates on the database.

```{r prepareupload}
data_for_upload <- prepare_transect_upload(agent_list = dq)
```

Then upload to the database. You will need an active database connection.

*Note: before uploading, confirm that your `ProjectShortName` is unique using `check_unique_project(ProjectShortName, con)` if you have not already done so.*

```{r upload, eval = FALSE}
con <- weda::weda_connect(password = keyring::key_get(service = "ari-dev-weda-psql-01",
                                                      username = "psql_user"))

upload_transect_data(
  con              = con,
  data_list        = data_for_upload,
  uploadername     = "Firstname Surname",
  schema           = "transect"
)
```

The function appends data to the three raw tables (`raw_transect_records`, `raw_transects`, `raw_project_information`). The upload may take a few minutes. Leave your R session running and be patient.

To upload to the development schema first (recommended when testing a new project for the first time):

```{r uploaddev, eval = FALSE}
upload_transect_data(
  con          = con,
  data_list    = data_for_upload,
  uploadername = "Firstname Surname",
  schema       = "transect_dev"
)
```

# Accessing uploaded data

Once uploaded, use the curated view functions to query the most recent records without duplicates:

```{r accessdata, eval = FALSE}
# Curated records (SQL query, collected on demand)
transect_records_curated_view(con, return_data = TRUE)
transect_curated_view(con, return_data = TRUE)
transect_project_curated_view(con, return_data = TRUE)

# Processed presence/absence (all species)
processed_transect_presence_absence(con, return_data = TRUE)

# Presence/absence for a specific species
processed_transect_presence_absence(con,
                                    return_data = TRUE,
                                    species     = "Greater Glider")
```

See `vignette('data-download')` for more details on working with database data in R.

# Appendix {#appendix}

A data dictionary is provided in this package (`data(data_dictionary)`) and also available in the `data_dictionary` schema on the database. Below is the data dictionary for the `transect` schema:

```{r datadict}
data_dictionary %>%
  filter(schema == "transect") %>%
  select(table_name, table_description, column_name, column_class, column_description) %>%
  kbl() %>%
  kable_styling(c("condensed"), full_width = FALSE) %>%
  collapse_rows(1:3, valign = "top") %>%
  scroll_box(width = "100%", height = "1000px")
```
